---
title: "FP4"
author: "Gregory Forsberg"
date: "2025-11-14"
output: html_document
---

#loading packages
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggrepel)
library(scales)
library(treemapify)
library(ggpubr)
library(forcats)
library(sf)
library(countrycode)
library(rnaturalearth)
library(rnaturalearthdata)
library(ineq)
```

#loading in data
- ClinicalTrials.gov — Global Interventional Cancer Trials: Central repository for interventional and observational trials; includes study design, enrollment, and outcomes
- IHME Global Burden of Disease (GBD) — Global Cancer DALYs: Provides global epidemiological data on mortality, morbidity, and risk factors across countries and disease categories
```{r}
totalcancer <- read_csv("../Data/IHME-GBD_2021_DATA-eae6be09-1.csv")
clean_scrape_data <- read.csv("../Data/ctg_clean.csv")
world <- ne_countries(scale = "medium", returnclass = "sf")
cancerdaly <- read_csv("../Data/IHME-GBD_2021_DATA-12473b24-1.csv")
```
#Gregory
```{r}
total_by_year <- totalcancer %>%
  filter(metric == "Number", measure == "Deaths") %>%  
  group_by(year, cause) %>%
  summarise(total = sum(val, na.rm = TRUE), .groups = "drop")

top_causes <- total_by_year %>%
  group_by(cause) %>%
  summarise(overall_total = sum(total, na.rm = TRUE), .groups = "drop") %>%
  slice_max(overall_total, n = 10) %>%   
  pull(cause)

top_cancers_only <- total_by_year %>%
  filter(cause %in% top_causes)

ggplot(top_cancers_only, aes(x = year, y = total, color = cause)) +
  geom_line(size = 1) +
  labs(title = "Cancer Types by Total Deaths Over Time (Top 10)",
       x = "Year", y = "Total Deaths", color = "Cancer Type") +
    scale_y_continuous(labels = comma) +
  theme_minimal()

ct_count <- clean_scrape_data %>% 
  group_by(Condition_clean) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count)) %>% 
  slice_head(n = 10)

ggplot(ct_count,aes(reorder(Condition_clean, count), count))+
  geom_col()+
  theme_minimal()+
  labs(x = "Clinical trial targeted Condition")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) 

latest_year <- max(totalcancer$year, na.rm = TRUE)

global_latest <- totalcancer %>%
  filter(location == "Global", metric == "Number", measure == "Deaths", age == "All ages", sex == "Both", year == latest_year) %>%
  group_by(cause) %>%
  summarise(total_deaths = sum(val, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_deaths)) %>%
  slice_head(n = 15)

ggplot(global_latest, aes(x = reorder(cause, total_deaths), y = total_deaths)) +
  geom_col(fill = "blue") +
  coord_flip() +
  labs(title = paste("Top 15 Global Cancer Deaths in", latest_year), x = "Cancer Type", y = "Total Deaths") +
  theme_minimal()

global_trend <- totalcancer %>%
  filter(location == "Global", metric == "Number", measure == "Deaths", age == "All ages", sex == "Both") %>%
  group_by(year) %>%
  summarise(total_deaths = sum(val, na.rm = TRUE), .groups = "drop")

ggplot(global_trend, aes(x = year, y = total_deaths)) +
  geom_line(size = 1.2, color = "red") +
  geom_point(color = "darkred") +
  labs(title = "Total Global Cancer Deaths Over Time", x = "Year", y = "Total Deaths") +
  theme_minimal()

rate_dalys_trend <- cancerdaly %>%
  filter(location_name == "America",
         measure_name == "DALYs (Disability-Adjusted Life Years)",
         metric_name == "Rate", #100,000?
         sex_name == "Both",
         age_name == "All ages",
         cause_name != "Neoplasms") %>%
  group_by(year) %>%
  summarise(avg_daly_rate = mean(val, na.rm = TRUE), .groups = "drop")

ggplot(rate_dalys_trend, aes(x = year, y = avg_daly_rate)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(color = "green") +
  labs(title = "Cancer DALY Rate per 100k (America)",
       x = "Year", y = "DALYs per 100,000") +
  theme_minimal()
```
1) This visualization shows the death rate (per 100,000) by cancer type from 1980 to 2021 for the top ten causes of cancer deaths. The x-axis displays the year, and the y-axis shows the total death counts, with each colored line representing the different cancer types. This visualization address one of our research questions being: How have global cancer death rates changed over time? Understanding these important trends help drive public health and policy efforts and future medical research in treating cancer.

The highest death rate during this period is for tracheal, bronchus, and lung cancer which peaked in about the 1990s and has since slowly declined. This could be a result of a decrease in smoking among younger populations and/or advancement made in medicine to treat this type of cancer. Stomach cancer shows a significant and consistent decline, while most other cancer types remained stable or trend downward, though more gradually. These patterns reflect a variety of different factors such as medical advancements, changes in societal habits, and preventative care initiatives.

2) Graph looks at the raw totals of clinical trials based off the types of neoplasm that they target.

3) Graph looks at the global cancer death rates for top 15 cancer deaths in 2021.

4) Graph looks at the total global cancer deaths over time from 2012 to 2021.

5) Graph DALY rate per 100k from 2012 to 2021 within america for cancer.

#Khalessi
Has done a great amount of work on our llm to scrape the data from the clinical gov website. The code for the llm is in the code file titled "Clean.RMD". It isn't included in this rmd so that we don't re-run the llm. 
```{r}
ctg <- read.csv("../Data/ctg_clean.csv")
gbd <- read.csv("../Data/gbd_death.csv")
```

```{r}
trial_by_cancer <- ctg %>%
  count(Condition_clean, name = "trial_count")

gbd_latest <- gbd %>%
  filter(
    location == "Global",
    sex == "Both",
    metric == "Number",     
    measure == "Deaths",
    year == max(year)
  ) %>%
  select(cause, death_number = val)

compare_df <- trial_by_cancer %>%
  inner_join(gbd_latest, by = c("Condition_clean" = "cause")) %>%
  mutate(
    total_trials      = sum(trial_count, na.rm = TRUE),
    total_deaths      = sum(death_number, na.rm = TRUE),
    trials_share      = trial_count / total_trials,
    death_share       = death_number / total_deaths,
    ratio             = trials_share / death_share,
    status            = if_else(ratio < 1, "Under-studied", "Over-studied")
  )
```

```{r}
p1 <- ggplot(compare_df, aes(x = death_share, y = trials_share, label = Condition_clean)) +
  geom_point(aes(size = trial_count, color = status), alpha = 0.85) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_text_repel(max.overlaps = 12, size = 3) +
  scale_x_log10(labels = percent_format(accuracy = 0.01)) +
  scale_y_log10(labels = percent_format(accuracy = 0.01)) +
  scale_color_manual(
    values = c("Under-studied" = "#E74C3C", "Over-studied" = "#2E8B57"),
    name = "Research Attention"
  ) +
  scale_size_continuous(name = "Trial count") +
  labs(
    title = "Clinical Trial Share vs. Death Share by Cancer Type",
    subtitle = "Using global number of deaths (GBD)  — log-scaled shares",
    x = "Death share (proportion of total cancer deaths)",
    y = "Trial share (proportion of total cancer trials)"
  ) +
  theme_minimal(base_size = 13)
p1
```
```{r}
library(lubridate)

p2 <- ctg %>%
  mutate(Year = year(ym(str_sub(`Start.Date`, 1, 7)))) %>%
  count(Year) %>%
  ggplot(aes(x = Year, y = n)) +
  geom_line(color = "#E69F00", linewidth = 1) +
  scale_x_continuous(
    limits = c(1970, 2025),        # 👈 force scale to end at 2025
    breaks = seq(1970, 2025, by = 5)
  ) +
  labs(
    title = "Growth of Cancer Clinical Trials (1970–2025)",
    x = "Year of trial start",
    y = "Number of new trials"
  ) +
  theme_minimal(base_size = 13)

p2
```


#Scarlett
```{r}
#data
IHME_data <- read_csv("../Data/IHME_data.csv")
world <- ne_countries(scale = "medium", returnclass = "sf")


```

```{r}
IHME_data <- IHME_data %>%
  mutate(location = recode(location,
    "Viet Nam"                               = "Vietnam",
    "Venezuela(Bolivarian Republic of)"      = "Venezuela",
    "Syrian Arab Republic"                   = "Syria",
    "Republic of Korea"   = "South Korea",
    "Democratic People's Republic of Korea"  = "North Korea",
    "Russian Federation"                     = "Russia",
    "Taiwan(Province of China)"              = "Taiwan",
    "United Republic of Tanzania"            = "Tanzania",
    "Lao People's Democratic Republic"       = "Laos",
    "Iran (Islamic Republic of)"             = "Iran",
    "Republic of Moldova"                    = "Moldova",
    "Eswatini"                               = "Swaziland",
    "Bolivia (Plurinational State of)"       = "Bolivia",
    "Democratic Republic of the Congo" = "Congo (Kinshasa)",
    "Republic of the Congo" = "Congo (Brazzaville)",
    "Micronesia (Federated States of)"   = "Micronesia",
    "Marshall Islands"                   = "Marshall Is.",
    "Northern Mariana Islands"           = "N. Mariana Is.",
    "United States Virgin Islands"       = "U.S. Virgin Is.",
    "Antigua and Barbuda"                = "Antigua",
    "Saint Kitts and Nevis"              = "Saint Kitts",
    "Dominican Republic"                 = "Dominican Rep.",
    "Sao Tome and Principe"              = "Sao Tome & Principe",
    "Saint Vincent and the Grenadines"   = "Saint Vincent",
    "Cook Islands"                       = "Cook Is.",
    "Solomon Islands"                    = "Solomon Is.",
    "Equatorial Guinea"                  = "Eq. Guinea",
    "Bosnia and Herzegovina"             = "Bosnia",
    "Brunei Darussalam"                  = "Brunei",
    "South Sudan"                        = "S. Sudan",
    "Central African Republic"           = "Central African Rep.",
    "Tokelau"                            = "Tokelau"   # may not exist in world map

  ))

```


```{r, eval=FALSE}
top4<- IHME_data%>%
  group_by(cause)%>%
  summarise(val = mean(val))%>%
  arrange(desc(val))
```

**Visualization**
```{r}
world <- ne_countries(returnclass = "sf")

focus_cause <- "Stomach cancer"

focus_cause2 <- "Tracheal, bronchus, and lung cancer"

focus_cause3 <- "Breast cancer"

focus_cause4 <- "Colon and rectum cancer"

ihme_map_stomach <- IHME_data %>%
  filter(cause == focus_cause, 
         measure == "Deaths", 
         metric == "Rate") %>%
  group_by(location) %>%
  summarise(val = mean(val, na.rm = TRUE), .groups="drop")


mapdata <- world %>%
  left_join(ihme_map_stomach, by = c("name" = "location"))

Liver_cancer <- ggplot(mapdata) +
  geom_sf(aes(fill = val), color = "white", size = 0.2) +
  labs(caption = "Age-standardized death rate for Stomach cancer",
       fill = "Rate per 100k") +
  theme_classic()+
  theme(
    plot.caption = element_text(hjust = 0.5, size = 14),  
    plot.caption.position = "plot")+
  scale_fill_stepsn(
  colours = c("#1a9850", "#fee08b", "#d73039"),
  breaks = c(5, 10, 15, 20, 30, 40, 60), 
  na.value = "grey90",
  name = "Rate per 100k"
)

```





```{r}
ihme_map_lung <- IHME_data %>%
  filter(cause == focus_cause2, 
         measure == "Deaths", 
         metric == "Rate") %>%
  group_by(location) %>%
  summarise(val = mean(val, na.rm = TRUE), .groups="drop")


mapdata2 <- world %>%
  left_join(ihme_map_lung, by = c("name" = "location"))

Lung_cancer <- ggplot(mapdata2) +
  geom_sf(aes(fill = val), color = "white", size = 0.2) +
  labs(caption = "Age-standardized death rate for Tracheal, bronchus, and Lung cancer",
       fill = "Rate per 100k") +
  theme_classic()+
  theme(
    plot.caption = element_text(hjust = 0.5, size = 14),  
    plot.caption.position = "plot")+
  scale_fill_stepsn(
  colours = c("#1a9850", "#fee08b", "#d73039"),
  breaks = c(5, 10, 15, 20, 30, 40, 60), 
  na.value = "grey90",
  name = "Rate per 100k"
)
```

```{r}
ihme_map_breast <- IHME_data %>%
  filter(cause == focus_cause3, 
         measure == "Deaths", 
         metric == "Rate") %>%
  group_by(location) %>%
  summarise(val = mean(val, na.rm = TRUE), .groups="drop")


mapdata3 <- world %>%
  left_join(ihme_map_breast, by = c("name" = "location"))

Breast_cancer <- ggplot(mapdata3) +
  geom_sf(aes(fill = val), color = "white", size = 0.2) +
  labs(caption = "Age-standardized death rate for Breast cancer",
       fill = "Rate per 100k") +
  theme_classic()+
  theme(
    plot.caption = element_text(hjust = 0.5, size = 14),  
    plot.caption.position = "plot")+
  scale_fill_stepsn(
  colours = c("#1a9850", "#fee08b", "#d73039"),
  breaks = c(5, 10, 15, 20, 30, 40, 60), 
  na.value = "grey90",
  name = "Rate per 100k"
)
```

```{r}
ihme_map_colon <- IHME_data %>%
  filter(cause == focus_cause4, 
         measure == "Deaths", 
         metric == "Rate") %>%
  group_by(location) %>%
  summarise(val = mean(val, na.rm = TRUE), .groups="drop")


mapdata4 <- world %>%
  left_join(ihme_map_colon, by = c("name" = "location"))

Colon_cancer <- ggplot(mapdata4) +
  geom_sf(aes(fill = val), color = "white", size = 0.2) +
  labs(caption = "Age-standardized death rate for Colon and Rectum cancer",
       fill = "Rate per 100k") +
  theme_classic() +
  theme(
    plot.caption = element_text(hjust = 0.5, size = 14),  
    plot.caption.position = "plot")+
  scale_fill_stepsn(
  colours = c("#1a9850", "#fee08b", "#d73039"),
  breaks = c(5, 10, 15, 20, 30, 40, 60), 
  na.value = "grey90",
  name = "Rate per 100k"
)
```

**Final Visual**
```{r fig.alt="Four world maps showing age-standardized cancer mortality rates for stomach, lung, colorectal, and breast cancer. Data from the Global Burden of Disease study.", fig.width=13, fig.height=8}
combined <- ggarrange(Liver_cancer, 
                              Lung_cancer, 
                              Colon_cancer, 
                              Breast_cancer, 
                              ncol = 2, nrow = 2, 
                              widths = c(2, 2),     
                              heights = c(2, 2)
                              )

final_cancer_map <- annotate_figure(
  combined,
  top = text_grob("Global Age-standardized Cancer Mortality Rates", 
                  size = 18, hjust = 0.5)
)

final_cancer_map
```


