---
title: "Clinical Trials vs Disease Burden Final report"
author: "Gregory, Khalessi, Scarlett"
date: "2025-12-12"
output: html_document
---

## Topic Introduction
Project Title: Clinical Trials vs Disease Burden (Neoplasms)
Gregory, Khalessi, Scarlett

For this project, we are looking into how clinical trials around the world are addressing the global burden of disease, focusing on cancer/neoplasms. 

## Motivations and Research Question
Our focus for this final project is centered around looking into how the global burden of disease is being addressed by clinical trials around the world. With cancer being a world-wide health issue, the estimated number of cancer cases world-wide is expected to grow to around 35 million by 2050. Which such a high number, it is paramount to human health that cancer research continues to discover cancer treatments for all types. Because challenging complex medical diseases is extremely expensive, certain types of cancer research receives more funding either to do better publicity or larger effect group. This can lead to an imbalance of research for the "understudied" cancer types which still remain extremely deadly. As a result, our research questions aim to directly address this issues of, How does clinical research around the world address known disease burdens? In other words, we are asking, how do the number/type of clinical trials address the current need for treatments among known diseases that effect many people. 

## Data sources
1) Global Burden of Disease Study (GBD)

This annual study is produced by the Institute for Health Metrics and Evaluation and combines millions of health statistics from organizations around the world into a large, centralized report. For this project, the dataset provides global information on cancer mortality and disease burden across countries using mortality rates and disability-adjusted life years (DALYs) over time, from 1980 to the present. The data were obtained in CSV format through the GBD interactive dashboard, where query filters were used to select the desired statistics and years. The dataset required minimal additional cleaning.

2) ClinicalTrials.gov

ClinicalTrials.gov is a publicly accessible registry of clinical trials maintained by the National Institutes of Health (NIH). It serves as a centralized database for both interventional and observational clinical studies conducted worldwide and includes information on study location, disease focus, enrollment size, study phase, and trial status. The data from this source were highly unstructured for the purposes of our analysis, which required the use of a large language model (LLM) to standardize trial types based on study titles.

3) WHO International Clinical Trials Registry Platform (ICTRP)

The World Health Organization’s International Clinical Trials Registry Platform (ICTRP) stores clinical trial records from national and regional registries around the world. Data from this registry were used as a supplement to the NIH ClinicalTrials.gov dataset to improve coverage and transparency of clinical trials across both databases.


```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = FALSE, message = FALSE, warning = FALSE,
fig.align = "center", fig.width = 11, fig.height = 6
)
```


```{r}
library(ellmer);
library(tidyverse)
library(furrr)
library(kableExtra);
library(jsonlite)
library(dplyr)
library(readr)
library(httr2)
library(purrr)
library(tibble)
library(stringr)
library(scales)
library(ggpubr)
library(ggrepel)
library(forcats) 
library(ggplot2)
library(sf)
library(countrycode)
library(rnaturalearth)
library(rnaturalearthdata)
```


```{r pressure, echo=FALSE}
gbd_mertic <- read_csv("../Data/gbd_mertic.csv")
clinic <- read_csv("../Data/clinic2.0.csv")
```
## Research Question 1

### How do clinical trials address the global disease burden of cancer?


```{r echo=TRUE}
clinic_by_cancer <- clinic %>%
  filter(!is.na(Condition_clean)) %>%
  group_by(Condition_clean) %>%
  summarise(
    n_trials     = n_distinct(TrialID),
    total_target = sum(target_total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(trial_share = n_trials / sum(n_trials))
make_joined_df <- function(measure_name) {
  gbd_latest <- gbd_mertic %>%
    filter(
      measure  == measure_name,
      location == "Global",
      sex      == "Both",
      age      == "All ages",
      metric   == "Number"
    ) %>%
    group_by(cause) %>%
    filter(year == max(year)) %>%
    ungroup() %>%
    select(cause, val)

  gbd_latest <- gbd_latest %>%
    mutate(burden_share = val / sum(val))

  clinic_by_cancer %>%
    inner_join(gbd_latest, by = c("Condition_clean" = "cause")) %>%
    mutate(
      over_rep = trial_share > burden_share   # TRUE = green, FALSE = red
    )
}

make_plot <- function(measure_name, title_suffix) {
  df <- make_joined_df(measure_name)

  ggplot(df, aes(x = burden_share, y = trial_share)) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "grey50") +
    geom_point(aes(color = over_rep, size = n_trials), alpha = 0.8) +
    geom_text_repel(
      aes(label = Condition_clean),
      size = 3,
      max.overlaps = Inf,   
      box.padding = 0.3,
      point.padding = 0.2,
      segment.color = "grey60"
    ) +
    scale_x_log10(labels = scales::percent_format(accuracy = 0.01)) +
    scale_y_log10(labels = scales::percent_format(accuracy = 0.01)) +
    scale_color_manual(
      values = c("FALSE" = "red", "TRUE" = "darkgreen"),
      labels = c("Underrepresented", "Overrepresented"),
      name   = NULL
    ) +
    labs(
      x = paste0(title_suffix, " share"),
      y = "Clinical trial share",
      title = paste("Clinical trial share vs", title_suffix, "by cancer type")
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )
}

p_dalys  <- make_plot("DALYs (Disability-Adjusted Life Years)", "DALY")

p_dalys

```


```{r echo=TRUE}
gbd_cn <- read_csv("../Data/gbd_cn.csv")

clinic_cn <- read_csv("../Data/clinic_cn.csv")

```

```{r echo=TRUE}

ratio_df <- gbd_cn %>%
  inner_join(clinic_cn, by = c("country_std" = "country_std")) %>%
  mutate(
    total_trials      = sum(n_trials, na.rm = TRUE),
    total_daly      = sum(dalys, na.rm = TRUE),
    trials_ratio      = n_trials / total_trials,
    daly_ratio       = dalys / total_daly,
    ratio             = trials_ratio / daly_ratio
    ) %>%
  
  select(country_std, dalys, n_trials, total_trials, total_daly, daly_ratio, trials_ratio, ratio)
  
```

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

ratio_df <- ratio_df %>%
  mutate(
    country_std = recode(country_std,
      "United States"               = "United States of America",
      "United Republic Of Tanzania" = "Tanzania",
      "Viet Nam"                    = "Vietnam",
      "Russian Federation"          = "Russia",
      "Republic Of Korea"           = "South Korea",                  
      "Western Sahara" = "W. Sahara",
      "Laos" = "Lao PDR",
      "Palestine" = "Palestinian Territories",
      "Eq. Guinea" = "Equatorial Guinea",
      "N. Cyprus" = "Northern Cyprus",
      "Côte d'Ivoire" = "Ivory Coast",
      "Dem. Rep. Congo" = "Democratic Republic of the Congo",
      "Congo" = "Republic of the Congo",
      "Macao" = "Macau",
      "Indian Ocean Ter." = "British Indian Ocean Territory",
      "Heard I. and McDonald Is." = "Heard Island and McDonald Islands",
      "Ashmore and Cartier Is." = "Ashmore and Cartier Islands",
      "Antigua and Barb." = "Antigua and Barbuda", "Fiji" = "Fiji",
    "Papua New Guinea" = "Papua New Guinea",
    "Democratic Republic of the Congo" = "Dem. Rep. Congo",
  
    "Falkland Islands" = "Falkland Is.",
    "French Southern Territories" = "Fr. S. Antarctic Lands",
    "Bolivia (Plurinational State of)" = "Bolivia",
  
    "Venezuela (Bolivarian Republic of)" = "Venezuela",
   
    "Cote d'Ivoire" = "Côte d'Ivoire",
  
  
    "Central African Republic" = "Central African Rep.",
    "Republic of the Congo" = "Congo",
    "Equatorial Guinea" = "Eq. Guinea",
    "Eswatini" = "eSwatini",
    "State of Palestine" = "Palestine",
    
    "Lao People's Democratic Republic" = "Laos",
  
    "Democratic People’s Republic of Korea" = "North Korea",
  
 
    "Iran (Islamic Republic of)" = "Iran",
    "Syrian Arab Republic" = "Syria",
    "Republic of Moldova" = "Moldova",
    "Türkiye" = "Turkey",
    "New Caledonia" = "New Caledonia",
    "Solomon Islands" = "Solomon Is.",
    
    "Brunei Darussalam" = "Brunei",
  

    "Northern Cyprus" = "N. Cyprus",
   

    "South Sudan" = "S. Sudan",
    "Bosnia and Herzegovina" = "Bosnia and Herz.",
    "Republic Of Korea"      = "South Korea",
    "Republic of Korea"      = "South Korea", 
    "Dominican Republic" = "Dominican Rep."
    )
  )

```

```{r echo=TRUE}
world_sf <- ne_countries(returnclass = "sf")


map_df <- world_sf %>%
  left_join(ratio_df, by = c("name" = "country_std")) %>%   # change keys if needed
  mutate(ratio_cat = cut(
    ratio,
    breaks = c(-Inf, 0.25, 0.5, 1, 2, Inf),
    labels = c("<0.25", "0.25–0.5", "0.5–1", "1–2", ">2")
  ))

# Define a diverging palette that highlights >1 clearly
ratio_colors <- c(
   "<0.25" = "#d7191c",
    "0.25–0.5" = "#fdae61",
    "0.5–1" = "#ffffbf",
    "1–2" = "#91B5F2",
    ">2" = "#2b83ba",
    "NA" = "grey80"
)

ratio_world_map <- ggplot(map_df) +
  geom_sf(aes(fill = ratio_cat), color = "white", size = 0.2) +
  coord_sf(expand = FALSE) +
  scale_fill_manual(
    values = ratio_colors,
    drop = FALSE,
    name = "Trial share / Burden share"
  ) +
  labs(
    title = "Clinical Trial Activity vs. Disease Burden",
    caption = "Countries with ratios > 1 have more clinical trials\nrelative to their share of disease burden."
  ) +
  theme_void() +
  theme(
    axis.title  = element_blank(),
    axis.text   = element_blank(),
    axis.ticks  = element_blank(),
    plot.caption = element_text(hjust = 0.5, size = 8),
    plot.caption.position = "plot",
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 9)
  )


ratio_world_map

```


```{r}
daly_ratio_df <- gbd_cn %>%
  mutate(
    total_daly      = sum(dalys, na.rm = TRUE),
    daly_ratio       = dalys / total_daly,
    ) %>%
  
  select(country_std, dalys, total_daly, daly_ratio)

daly_ratio_df_clean <- daly_ratio_df %>%
  mutate(country_std = recode(country_std,
      "United States"               = "United States of America",
      "United Republic Of Tanzania" = "Tanzania",
      "Viet Nam"                    = "Vietnam",
      "Russian Federation"          = "Russia",
      "Republic Of Korea"           = "South Korea",                  
      "Western Sahara" = "W. Sahara",
      "Laos" = "Lao PDR",
      "Eq. Guinea" = "Equatorial Guinea",
      "N. Cyprus" = "Northern Cyprus",

  
      "Macao" = "Macau",
      "Indian Ocean Ter." = "British Indian Ocean Territory",
      "Heard I. and McDonald Is." = "Heard Island and McDonald Islands",
      "Ashmore and Cartier Is." = "Ashmore and Cartier Islands",
      "Antigua and Barb." = "Antigua and Barbuda", "Fiji" = "Fiji",
 
   
    "Falkland Islands" = "Falkland Is.",
    "French Southern Territories" = "Fr. S. Antarctic Lands",
    "Bolivia (Plurinational State of)" = "Bolivia",
  
    "Venezuela (Bolivarian Republic of)" = "Venezuela",
  
  
  
    "Central African Republic" = "Central African Rep.",
  
    "Equatorial Guinea" = "Eq. Guinea",
    "Eswatini" = "eSwatini",
    "State of Palestine" = "Palestine",
    
    "Lao People's Democratic Republic" = "Laos",
  
  
 
    "Iran (Islamic Republic of)" = "Iran",
    "Syrian Arab Republic" = "Syria",

    "Türkiye" = "Turkey",
    "New Caledonia" = "New Caledonia",
    "Solomon Islands" = "Solomon Is.",
    
    "Brunei Darussalam" = "Brunei",
  

    "Northern Cyprus" = "N. Cyprus",
   

    "South Sudan" = "S. Sudan",
   
    "Republic Of Korea"      = "South Korea",
    "Republic of Korea"      = "South Korea", 
    "Dominican Republic" = "Dominican Rep.", 
    
      "Democratic Republic Of The Congo"      = "Dem. Rep. Congo",
      "Republic of the Congo"                 = "Congo",
      "Bosnia And Herzegovina"                = "Bosnia and Herz.",
      "Iran (Islamic Republic Of)"            = "Iran",
      "Republic Of Moldova"                   = "Moldova",
      "Sao Tome And Principe"                 = "Sao Tome and Principe",
      "Trinidad And Tobago"                   = "Trinidad and Tobago",
      "Democratic People's Republic Of Korea" = "North Korea",
      "Bolivia (Plurinational State Of)"      = "Bolivia",
      "Venezuela (Bolivarian Republic Of)"    = "Venezuela",
      "Côte D'ivoire"                         = "Côte d'Ivoire",
      "Cabo Verde"                            = "Cape Verde",
      "Micronesia (Federated States Of)"      = "Micronesia",
      "Palestinian Territories"               = "Palestine"


    
  ))

burden_map_df <- world %>%
  left_join(daly_ratio_df_clean, by = c("name" = "country_std"))
```

```{r echo=TRUE}

burden_world_map <- ggplot(burden_map_df) +
  geom_sf(aes(fill = daly_ratio), color = "white", size = 0.2) +
  coord_sf(expand = FALSE) +
  scale_fill_gradientn(
  colors = c("#e0ecf4", "#9ebcda", "#8c6bb1", "#88419d", "#6e016b"),
  trans = "sqrt",
  labels = percent_format(accuracy = 0.1),
  name = "Share of global\nDALY burden"
  ) +
  labs(
    title   = "Global Distribution of Disease Burden (DALYs)",
    caption = "Darker colors indicate countries with a higher share of global DALY burden."
  ) +
  theme_void() +
  theme(
    axis.title  = element_blank(),
    axis.text   = element_blank(),
    axis.ticks  = element_blank(),
    plot.caption = element_text(hjust = 0.5, size = 8),
    plot.caption.position = "plot",
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 9)
  )

burden_world_map

```

**Result (DALY-based alignment).**

Clinical trial activity does **not uniformly align** with the global burden of cancer as measured by Disability-Adjusted Life Years (DALYs). The DALY-based trial–burden scatterplot shows substantial dispersion around the line of proportionality, indicating that many cancer types receive either more or fewer trials than expected given their contribution to global cancer DALYs.

While some high-burden cancers—such as tracheal, bronchus, and lung cancer, breast cancer, and colorectal cancer —are relatively well aligned with DALY burden, several cancers with substantial DALY shares, including liver cancer, stomach cancer, and esophageal cancer, are underrepresented in clinical trials.

These findings indicate that global cancer clinical research prioritization is **only partially driven by disease burden**. Factors such as therapeutic feasibility, availability of molecular targets, existing research infrastructure, and pharmaceutical investment likely play a significant role in shaping where trials are conducted. As a result, DALYs alone do not fully predict where clinical research activity is concentrated.

Overall, clinical trials address global cancer burden **imperfectly**: while some major contributors to DALYs receive substantial research attention, notable gaps persist for several high-burden cancers.

## Research Question 2
### How does socio-demographic index influence clinical trials and the burden of disease?

```{r echo=TRUE}
SDI <- read_csv("../Data/SDI.csv")
gbd_country <- read_csv("../Data/gbd_country.csv")
```

```{r}
SDI_clean <- SDI %>%
  rename(
    country_sdi = `Location Name`,
    SDI_value = `2023 SDI Index Value`,
    SDI_quintile = `SDI Quintile`
  ) %>%
  mutate(country_std = str_to_title(country_sdi))
```

```{r}
trials_country <- clinic %>%
  filter(!is.na(Countries_clean)) %>%
  separate_rows(Countries_clean, sep = ";") %>%
  mutate(country = str_trim(Countries_clean),
         country_std = str_to_title(country)) %>%
  group_by(country_std) %>%
  summarise(
    n_trials = n_distinct(TrialID),
    total_target = sum(target_total, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
trials_with_sdi <- trials_country %>%
  left_join(SDI_clean, by = "country_std")
```


```{r}
SDI_missing <- SDI %>%
  filter(is.na(`SDI Quintile`) & !is.na(`2023 SDI Index Value`)) %>%
  select(`Location Name`, `2023 SDI Index Value`)
```

```{r}
standardize_country <- function(x) {
  x_utf8 <- iconv(x, from = "", to = "UTF-8")
  x_low <- tolower(str_trim(x_utf8))
  case_when(
    x_low %in% c("united states", "united states of america", "usa", "u.s.a.") ~ "United States",
    x_low %in% c("united kingdom", "uk", "u.k.")                               ~ "United Kingdom",
    x_low %in% c("russian federation", "russia")                               ~ "Russia",
    x_low %in% c("iran (islamic republic of)", "iran")                         ~ "Iran",
    x_low %in% c("viet nam", "vietnam")                                        ~ "Vietnam",
    TRUE                                                                       ~ str_to_title(x_low)
  )
}
```



```{r}
SDI_calculated <- SDI %>%
  rename(
    country_sdi  = `Location Name`,
    SDI_value    = `2023 SDI Index Value`,
    SDI_quintile = `SDI Quintile`
  ) %>%
  mutate(
    SDI_quintile_new = case_when(
      SDI_value < 0.46 ~ "Low SDI",
      SDI_value < 0.59 ~ "Low-middle SDI",
      SDI_value < 0.69 ~ "Middle SDI",
      SDI_value < 0.78 ~ "High-middle SDI",
      SDI_value >= 0.78 ~ "High SDI",
      TRUE ~ NA_character_
    ),
    SDI_quintile_final = ifelse(
      is.na(SDI_quintile),
      SDI_quintile_new,
      SDI_quintile
    ),
    country_std = standardize_country(country_sdi)
  ) %>%
  select(country_std, SDI_value, SDI_quintile_final)
```


```{r echo=TRUE}
gbd_country_burden <- gbd_country %>%
  filter(
    sex == "Both",
    age == "All ages",
    metric == "Number",
    measure == "DALYs (Disability-Adjusted Life Years)"
  ) %>%
  group_by(location) %>%
  filter(year == max(year)) %>%
  summarise(
    dalys = val,
    .groups = "drop"
  ) %>%
  mutate(country_std = str_to_title(location))

total_global_dalys <- sum(gbd_country_burden$dalys)

gbd_country_burden <- gbd_country_burden %>%
  mutate(burden_share = dalys / total_global_dalys)

gbd_country_burden <- gbd_country_burden %>%
  mutate(country_std = standardize_country(location))

trials_country <- clinic %>%
  filter(!is.na(Countries_clean)) %>%
  separate_rows(Countries_clean, sep = ";") %>%
  mutate(country_std = str_to_title(str_trim(Countries_clean))) %>%
  group_by(country_std) %>%
  summarise(
    n_trials = n_distinct(TrialID),
    total_target = sum(target_total, na.rm = TRUE),
    .groups = "drop"
  )

total_trials_global <- sum(trials_country$n_trials)

trials_country <- trials_country %>%
  mutate(trial_share = n_trials / total_trials_global)
```


```{r echo=TRUE}
SDI_clean_final <- SDI_calculated %>%
  mutate(country_std = str_to_title(country_std)) %>%
  select(country_std, SDI_value, SDI_quintile_final)

global_df <- gbd_country_burden %>%
  left_join(trials_country, by = "country_std") %>%
  left_join(SDI_clean_final, by = "country_std")

global_df <- global_df %>%
  mutate(
    mismatch = trial_share / burden_share,
    log2_mismatch = log2(mismatch)  # >0 = overrepresented; <0 = underrepresented
  )
```

```{r echo=TRUE}
SDIClinic <- ggplot(global_df,
       aes(x = SDI_value, y = log2_mismatch,
           color = SDI_quintile_final, label = country_std)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(max.overlaps = 18, size = 3) +
  labs(
    title = "Global Relationship Between SDI and Cancer Trial–Burden Mismatch",
    subtitle = "log2(trial_share / burden_share): values < 0 = under-represented; values > 0 = over-represented",
    x = "Socio-Demographic Index (SDI)",
    y = "Mismatch Index (log2 ratio)"
  ) +
  scale_color_brewer(palette = "Set1", na.value = "grey40") +
  theme_minimal(base_size = 12)
SDIClinic

```

**Result (SDI and trial–burden mismatch).**

There is a clear association between socio-demographic development (SDI) and the alignment between clinical trial activity and cancer burden. Countries with **higher SDI values** tend to have **positive trial–burden mismatch scores**, indicating that they host a larger share of clinical trials relative to their share of global cancer DALYs. In contrast, many **lower-SDI countries** exhibit negative mismatch values, reflecting **underrepresentation in clinical trials** despite substantial disease burden.

The SDI–mismatch scatterplot shows an upward trend: as SDI increases, countries are more likely to be overrepresented in clinical trials. High-SDI countries cluster above the zero line (log₂ mismatch > 0), whereas low- and middle-SDI countries more frequently fall below it.

This pattern suggests that clinical trial activity is strongly influenced by national capacity rather than burden alone. Higher SDI countries generally have greater research infrastructure, regulatory capacity, funding availability, and established academic–industry networks, making them more attractive sites for clinical trials. Meanwhile, lower SDI countries may face structural barriers to trial participation, including limited infrastructure, fewer trained personnel, and underrepresentation in international trial registries.

Socio-demographic development plays a critical role in shaping where cancer clinical trials occur. Even when disease burden is high, countries with lower SDI are less likely to host clinical trials, reinforcing global inequities in access to cancer research and innovation.

Several questions remain unanswered by this analysis. It is unclear whether cancers that are underrepresented in clinical trials experience slower treatment development or worse patient outcomes over time. Future research could examine how trial funding, pharmaceutical incentives, and regulatory barriers influence which cancers receive the most research attention. Additionally, it remains unknown whether expanding clinical trial access in lower-SDI countries would meaningfully reduce global cancer mortality or improve equity in treatment availability.

## Unanswered Questions/Limitations

Several unanswered questions remain from this analysis. It is unclear whether cancers that are underrepresented in clinical trials experience slower treatment development or worse patient outcomes over time. Future research could examine how trial funding and structural incentives influence which cancers receive the most research attention. In addition, it remains unclear whether expanding clinical trial efforts in lower-SDI countries would directly reduce disease burden or improve equity in treatment availability.

The limitations of this study include treating all clinical trials equally, even though they differ widely in size, phase, and impact. Another important limitation is that trial location may not fully reflect who benefits from the research, as many studies enroll patients across multiple countries. In addition, individuals who participate in clinical trials already have a cancer diagnosis, which may influence participation patterns and access. Clinical trial registries may also underrepresent certain regions due to inconsistent reporting, even when using both NIH and WHO datasets. Finally, DALY and mortality rates do not fully capture all aspects of disease severity or other clinical needs of patients.

## Project Contribution

Gregory was primarily responsible for much of the written components of the project and for analyzing overall trends and summary statistics. Khalessi focused on cleaning and standardizing the NIH ClinicalTrials.gov data, including the use of large language models to process unstructured trial information. Scarlett focused on geographic trends and the creation of visualizations, including mapping clinical trials and disease burden across regions.

## Ai Useage
AI was used to improve visualizations, check spelling and grammar, and assist with standardizing data within the llm.

